---
title: "Practical Machine Learning"
author: "Johan Kullingsjö"
date: "Friday, April 24, 2015"
output: html_document
---

## Background  

Six people have performed barbell lifts correctly and incorrectly in 5 different ways. About 160 features have been derived (including some metadata). We want to explore the dataset and try to predict how a person perform. To our help we have around 19000 classified records.

### Strategy  

1. Look through the variables  
  - Type of variables (Quantative, Qualitative ...) 
  - Distribution  
  - Missing Values  
  
2. Make a Principal Component Analysis 
  - First on raw variables
  - Then on transformed 
  
3. Look for outliers 
  - Remove outliers (whole records) from analysis  
  *Later, look at them separately*
  
4. Decide on strategy (ML-method)
  - Before looking at data we might do some guesses.  
    * Five different behaviours (A, B, C, D, E)
    * Six subjects pedro, adelmo, carlitos, charles, eurico, jeremy
    
    We can suspect to both see clustering for subjects and behaviour  
    It's likely to see some extreme values that can be seen as measure artefacts  
      
    We have almost 20 000 examples, but only six subjects  
    This means that if the suspicion is correct that the  
    subjects differ substantially from each other in a PCA  
    we will have difficulties to make a general model that would predict  
    well for new subjects. 
    
    To decide a final machinelearning algorithm is an iterative process so  
    where we finally ends depends on the performance on the different approaches we test.  
    
    How general should our model be?  
    In this particular case it seems that our test data concists of 
    data from same subjects as we have in our train data. If we don't  
    need to predict from data collected from an unknown person we should be   
    able to get pretty decent predictions. 
    
    Two models worth trying are 
      * PCA and KNN (one model for each subject)  
      * Random forrest like model  
        
    Both models will probably perform well on subjects in training data  
    But generalize very poorly to subjects not included in the model. 
    (Random forrest probably a bit better since it will likely find data from
    the subject most alike todo predictions, but still poor). Another benefit from 
    random forrest is that one model is enough and you don't need  
    the name/identifier of each person.  
        
      
    If we on the other hand wants to make a model that generalize well to new 
    subjects we need something different. Since we have so few subjects it will be 
    hard to know how well our model generalize to new persons (
    because we don't know how well our subjects describe the space of all body shapes )  
    Some metadata (height, weight, gender, age ...) would help (but that we don't have)  
      
    Two models worth trying are 
      * PCA and KNN (one model for each behaviour) , Measure distance to model 
      * Models that separates one behaviour from all others (create at least 
      one model foreach behaviour)  (something like SVM )
      
      If we are lucky these generalize well (though hard to know with only six subjects )  
      (and might be similar on both in sample 
      and out of sample predictions  compared to the less general model)
      
    
  5. Since the limitation is 5 plots I will only show the principal ideas and the end results in the plots.

## The analysis

LOAD LIBRARIES
```{r load_libraries, echo=TRUE, results='hide', message=FALSE  }
  # Load libraries
  # caret 
  library(dplyr)
  library(tidyr)
  library(pcaMethods)

```

Define functions for performing the suggested ML methods  

```{r define_functions, echo=TRUE, results='hide', message=FALSE  }
# my own KNN (change to other if time ...)
knn <- function(scores_orig, scores_pred, orig_data, k=10) {
  # predict behaviour class (classe column)
  
  dist <- apply(scores_orig, 1, FUN = (function(x) {sqrt(sum((scores_pred-x)**2))}))

  # rank the distvector
  # extract indices with rank <= k
  ind <- which(rank(dist)<=k)
  #print(ind)
  
  # select classe for those indices from orig data
  my_class <- orig_data[ind,] %>%
    select(classe)
  # look at distribution, choose class with highest density
  m <- data.frame(table(my_class))
  
  m <- m %>%
    tbl_df %>%
    arrange(desc(Freq))
  
  print(m)
  prediction <- m$my_class[1][[1]]
  
  return(as.character(prediction))
  
  # Todo as improvements
  # if equal, remove one until one is highest ...
  
  # we can also look at prior probabillities ... better do in other fct and ret table
  # from here
  
}


predict_multi <- function(data2predict_t, data2createModel, user, Pcs=6 , k=10) {
    # predict behaviour for multiple records
    model_data <- filter(data2createModel, outlier1!="strong",  user_name==user)
    res_pca <- pca(model_data, scale = "uv", center = TRUE,nPcs = Pcs)
# 
    my_load <- loadings(res_pca)
    varnames <- row.names(my_load)

    pred_data <-  data2predict_t[,varnames] #use these for prediction
  
  res_pred <- predict(object = res_pca, 
        newdata = pred_data,
        pcs = nP(res_pca), 
        pre = TRUE, 
        post = FALSE)
  
  apply(res_pred$scores,MARGIN = 1, FUN = (function(x) {
    x <- matrix(x, nrow = 1,dimnames = list(1,names(x)))
    knn(scores_orig=scores(res_pca), scores_pred=x, orig_data=model_data, k=k)
  }))
}  

  
```

```{r download_data, echo=FALSE, results='hide', message=FALSE }
# Make sure we have data
  #Training
if(!file.exists("pml-training.csv")){
  url1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"  
  download.file(url = url1,destfile = "pml-training.csv")
}

#Testing
if(!file.exists("pml-testing.csv")){
  url2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(url = url2, destfile = "pml-testing.csv")
}
  
```

Load data  
```{r load_data, echo=TRUE, results='hide', message=FALSE , cache=TRUE }
# load data, cache the result
train <- read.csv("pml-training.csv")
test <- read.csv("pml-testing.csv")
train <- tbl_df(train)
test <- tbl_df(test)
  
``` 

*Variable selection*
As a first selection on variables I check for missing values and only keep
measures that have few missing values and the metadata (identification, timestamp, behaviourclass ...). To look at the code, search for chunck name variable_selection  
In total 60 variables (including 7 meta variables) are kept  

```{r variable_selection, echo=FALSE, results='hide', message=FALSE  }
# Let's see how variables differ in test and training
base::setdiff(names(test), names(train)) # [1] "problem_id" (in test)
base::setdiff(names(train), names(test)) # [1] "classe" (in train)
glimpse(train)
df_classes <- sapply(train,FUN = class) %>%
  data.frame() 

# We have a list of variables and classes
names(df_classes) <- c("type")
df_classes <- df_classes  %>%
  mutate(varname = row.names(.)) %>%
  tbl_df %>% 
  select(varname, type) 

# summary(df_classes)
#    varname               type   
#  Length:160         factor :37  
#  Class :character   integer:35  
#  Mode  :character   numeric:88 
# To much to write out in the report but
# I'll look through and see if some of them should be coerced to other type
# etc

# as a start the first seven variables will not be used in the 
# evaluation. However, in the end, several of these are potentially useful
# 
# 1   integer                        X
# 2    factor                user_name
# 3   integer     raw_timestamp_part_1
# 4   integer     raw_timestamp_part_2
# 5    factor           cvtd_timestamp
# 6    factor               new_window
# 7   integer               num_window

# Source: local data frame [19,622 x 7]
# 
#     X user_name raw_timestamp_part_1 raw_timestamp_part_2   cvtd_timestamp new_window num_window
# 1   1  carlitos           1323084231               788290 05/12/2011 11:23         no         11
# 2   2  carlitos           1323084231               808298 05/12/2011 11:23         no         11


# Let's look at the factor type variables

df_factor <- filter(df_classes[-c(1:7),], type=="factor") %>%
  select(varname)

factor_train <- train[, df_factor[[1]]]

summary(factor_train)

# We can see that we have alot of missing values and #DIV/0! 
# since we only have around 300 or less values these are unlikely to
# have any predictive power
# they can be removed (and eventually be looked at separately)
# however, one variable we need was hidden in those, the classe variable
# we keep that one


df_factor <- df_factor[df_factor!="classe",]
keep <- base::setdiff(names(train), df_factor[[1]])
train2 <- train[, keep] # change later to train!
train2 %>%
  print

missing <- sapply(train2, FUN = function(x){sum(is.na(x)/length(x))*100}) %>%
  data.frame

names(missing) = c("percent_missing")
missing$varnames <- row.names(missing)
missing <- missing %>%
  tbl_df %>%
  select(varnames, percent_missing) %>%
  arrange(desc(percent_missing)) %>%
  filter(percent_missing <= 90) %>%
  print
  


# We are left with 60 variables (including metadata)
#keep <- base::setdiff(names(train2), missing[[1]])
train3 <- train2[, missing[,1][[1]]] 
train4 <- train3 %>%
  select(X:num_window,classe,roll_belt:magnet_forearm_z) %>%
  mutate(X=factor(X), user_name=factor(user_name), raw_timestamp_part_1=factor(raw_timestamp_part_1), raw_timestamp_part_2=factor(raw_timestamp_part_2),   cvtd_timestamp=factor(cvtd_timestamp), new_window=factor(new_window), num_window=factor(num_window))

```

*PCA overall*  
Principal Component Analysis, color on subjects. 
We have a clear separation between all subjects (except overlap in one case)  
When investigating higher components we discover one record is a severe outlier 
(Outlier can be identified with X = "5373", search code chunk pca_overall for code)  
We therefore add a column "outlier" to our dataset. This will be used to exclude this 
observation when we create our training models later on.  

```{r pca_overall, echo=FALSE, results='hide', message=FALSE , cache=TRUE,fig.keep='last'}
resALL <- pca(train4, scale = "uv", center = TRUE, nPcs = 6)
loadingALL <- loadings(resALL)
slplot(resALL,pcs = c(1,2))
slplot(resALL,pcs = c(1,2), scol=train4$user_name )

slplot(resALL,pcs = c(3,4), scol=train4$user_name )
# one strong outlier we need to remove
check_outliers <- train4 %>%
  select(starts_with("gyr"))
train4$outlier1 <- ""
train4$outlier1[train4$X == "5373"] <- "strong"
check_outliers$outlier[check_outliers$gyros_forearm_z>200] <- "strong"
check_outliers2 <- check_outliers %>% filter(outlier!="strong")
# boxplot not working!!!
# boxplot(x = select(check_outliers2, gyros_arm_z:gyros_forearm_z), col=train4$user_name)
#resALL2 <- pca(train4, scale = "uv", center = TRUE, nPcs = 6)
resALL2 <- pca(filter(train4, outlier1!="strong"), scale = "uv", center = TRUE, nPcs = 10)
slplot(resALL2,pcs = c(1,2), scol=train4$user_name[train4$outlier1!="strong"] )




``` 
  
```{r pca_behaviour_class, echo=FALSE, results='hide', message=FALSE ,fig.keep='none' }
resA <- pca(filter(train4, outlier1!="strong",  classe=="A"), scale = "uv", center = TRUE,nPcs = 6)
resB <- pca(filter(train4, outlier1!="strong", classe=="B"), scale = "uv", center = TRUE,nPcs = 6)
resC <- pca(filter(train4, outlier1!="strong", classe=="C"), scale = "uv", center = TRUE,nPcs = 6)
resD <- pca(filter(train4, outlier1!="strong", classe=="D"), scale = "uv", center = TRUE,nPcs = 6)
resE <- pca(filter(train4, outlier1!="strong", classe=="E"), scale = "uv", center = TRUE,nPcs = 6)

resALL <- pca(train, scale = "uv", center = TRUE)
slplot(resA, scol=filter(train4, outlier1!="strong",  classe=="A")
       %>%select(user_name)%>% (function(x) {x[[1]]}), pcs = c(5,6))

slplot(resB, scol=filter(train4, outlier1!="strong",  classe=="B")
       %>%select(user_name)%>% (function(x) {x[[1]]}), pcs = c(5,6))

slplot(resC, scol=filter(train4, outlier1!="strong",  classe=="C")
       %>%select(user_name)%>% (function(x) {x[[1]]}), pcs = c(1,2))

slplot(resD, scol=filter(train4, outlier1!="strong",  classe=="D")
       %>%select(user_name)%>% (function(x) {x[[1]]}), pcs = c(5,6))

slplot(resE, scol=filter(train4, outlier1!="strong",  classe=="E")
       %>%select(user_name)%>% (function(x) {x[[1]]}), pcs = c(5,6))



  
```
*PCA for subject Pedro,  
color by behaviour class (classe)*  

Only two first components shown. We can see slight tendencies to separation 
netween the different behaviours (code chunck PCA_subject)
```{r PCA_subject, echo=FALSE, results='hide', message=FALSE, fig.keep='last', warning=FALSE  }
  # Now let's look at one subject at a time
# adelmo carlitos charles eurico jeremy pedro
res_adelmo <- pca(filter(train4, outlier1!="strong",  user_name=="adelmo"), scale = "uv", center = TRUE,nPcs = 6)

res_carlitos <- pca(filter(train4, outlier1!="strong",  user_name=="carlitos"), scale = "uv", center = TRUE,nPcs = 6)
res_charles <- pca(filter(train4, outlier1!="strong",  user_name=="charles"), scale = "uv", center = TRUE,nPcs = 6)
res_eurico <- pca(filter(train4, outlier1!="strong",  user_name=="eurico"), scale = "uv", center = TRUE,nPcs = 6)
res_jeremy <- pca(filter(train4, outlier1!="strong",  user_name=="jeremy"), scale = "uv", center = TRUE,nPcs = 6)
res_pedro <- pca(filter(train4, outlier1!="strong",  user_name=="pedro"), scale = "uv", center = TRUE,nPcs = 6)


slplot(res_adelmo, scol=filter(train4, outlier1!="strong",  user_name=="adelmo")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(3,4))

slplot(res_carlitos, scol=filter(train4, outlier1!="strong",  user_name=="carlitos")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(3,4))
slplot(res_charles, scol=filter(train4, 
                                outlier1!="strong",  user_name=="charles")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(3,4))
slplot(res_eurico, scol=filter(train4, 
                               outlier1!="strong",  user_name=="eurico")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(3,4))
slplot(res_jeremy, scol=filter(train4, 
                               outlier1!="strong",  user_name=="jeremy")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(1,2))
slplot(res_pedro, scol=
         filter(train4, outlier1!="strong",  user_name=="pedro")
       %>%select(classe)%>% (function(x) {x[[1]]}), pcs = c(1,2))

  
```
  
# Prediction

<!---
http://www.bioconductor.org/packages/release/bioc/manuals/pcaMethods/man/pcaMethods.pdf

See page 50 for prediction
-->
```{r echo=FALSE, results='asis', message=FALSE , cache=TRUE }
# Running out of time, let's make things easy for us
# Our test data has names in it, thus, instead of trying a
# general solution we will make a prediction for each person
# in the traindata, this simplifies alot 
# (though won't work for new people)

# ## S3 method for class 'pcaRes'
# predict(object, newdata, pcs = nP(object), pre = TRUE,
# post = TRUE, ...)
# ## S4 method for signature 'pcaRes'
# predict(object, newdata, pcs = nP(object), pre = TRUE,
# post = TRUE, ...)
# res_pedro

# Pedro
# res_pca <- res_pedro
# my_load <- loadings(res_pca)
# varnames <- row.names(my_load)
# user="pedro"
# user="adelmo"
# user="carlitos"
# user="charles"
# user="eurico"
# user="jeremy"
users = c("pedro", "adelmo", "carlitos", "charles", "eurico", "jeremy")
#user1="pedro"

#data1 <-  filter(train4, outlier1!="strong",  user_name==user1)
#pred_data <- data[1,varnames]



# res_pred <- predict(object = res_pca, 
#         newdata = pred_data,
#         pcs = nP(res_pca), 
#         pre = TRUE, 
#         post = FALSE)



#knn(scores_orig=scores(res_pca), scores_pred=res_pred$scores, orig_data=data, k=10)
#data <-  filter(test,   user_name==user)

for(user in users) {
  data <-  filter(train4, outlier1!="strong",  user_name==user)
  pred_list <- predict_multi(data2predict_t = data[,], 
                           data2createModel = train4, 
                           user = user,Pcs = 8 , k=11 )
  # check correct classified
tt <- data %>% select(classe) %>% mutate(pred=pred_list, corr=(classe==pred))
sum(tt$corr)/nrow(tt)
print(paste("in sample error for PCA based on ", user, " is ", tt))
  
}

# data <-  filter(train4, outlier1!="strong",  user_name==user)
# #data <- data1
# 
# 
# pred_list <- predict_multi(data2predict_t = data[,], 
#                            data2createModel = train4, 
#                            user = user,Pcs = 8 , k=11 )
# 
# # check correct classified
# tt <- data %>% select(classe) %>% mutate(pred=pred_list, corr=(classe==pred))
# sum(tt$corr)/nrow(tt)
# 
# 

``` 


